---
layout: mathjax	
title: TriMatch
subtitle: Propensity Score Matching for Non-Binary Treatments
published: false
status: process
---
```{r, results='hide', echo=FALSE}
opts_chunk$set(comment=NA, fig.width=11, fig.path='trimatch/')
options(width=90)
```

#### Introduction


a. Propensity scores are estimated for three models:

$${ PS }_{ 1 }=e({ x }_{ { T }_{ 1 }C })=Pr(z=1|{ x }_{ { T }_{ 1 }C })$$
$${ PS }_{ 2 }=e({ x }_{ { T }_{ 2 }C })=Pr(z=1|{ x }_{ { T }_{ 2 }C })$$
$${ PS }_{ 2 }=e({ x }_{ { T }_{ 2 }{ T }_{ 1 } })=Pr(z=1|{ x }_{ { T }_{ 2 }{ T }_{ 1 } })$$

b. Match order is determined. The default behavior is to start with the larger of the two treatments, followed the second treatment, and lastly the control group. However, the match order is configurable vis-à-vis the `match.order` parameter. 

c. Three distance matrices are calculated, ${D}_{1}$, ${D}_{2}$, and ${D}_{3}$ corresponding to the propensity scores estimated in step a.

d. Distances greater than the caliper (.25 by default, see Rosenbaum & Rubin, 1985) are eliminated. The caliper is specified in standard units so .25 corresponds to one-quarter of one standard deviation. Therefore the threshold for each model in actual units will be different.

e. The first element from group one is selected and the `nmatch` smallest elements from group two with a distance less than the caliper are selected. Note that the default value for `nmatch` is `Inf` so that all elements within the caliper are considered. For large datasets this can have a substantial impact on execution time. Although setting `nmatch` to a smaller value will increase execution time during development, we recommend setting `nmatch` to `Inf` for final results.

f. For each element in group two selected in step e, the `nmatch` smallest elements from group three with a distance less thant he caliper are selected.

g. The distance between each element identified in step f and the starting element from group one is retrieved and those with distance less than the caliper are retained. All matched triplets are saved to a data frame.

h. Steps e to g are repeated for the remaining elements in group one.

i. The total distance for each matched triplet identified in steps e to h are calculated and the data frame is sorted on the total distance.

j. The `M` unique smallest matched triplets are returned.


#### Installation

The latest development version of `TriMatch` can be installed from [Github](http://github.com/jbryer/TriMatch) using the `devtools` package.

```{r install, eval=FALSE}
require(devtools)
install_github('TriMatch', 'jbryer')
```

Looad the required R packages and data frame.

```{r setup, results='hide', message=FALSE}
require(TriMatch)
require(ez)
```

#### Example One: New Student Outreach

The data in example one (`data(students)`) represent newly enrolled students in a distance education program. By the nature of the program all students are considered part-time. Moreover, enrollment in the institution does not necessarily mean students are making active progress towards degree completion. To address this issue the institution began an outreach program whereby academic advisors would regularly contact new students within the first six months of enrollment until either six months have passed, or the student enrolled in some credit earning activity (generally a course or an examination for credit). Since randomization to receive the outreach was not possible, a comparison group was defined as students who enrolled six months prior to the start of the outreach. The treatment group is identified as students who enrolled six months after the start of the outreach and who received at least one academic advisor contact. 

Covariates for estimating propensity scores were retrieved from the student information system. The dependent variable of interest is the number of credits attempted within the first seven months of enrollment.

During the implementation phase it was identified that the outreach being conducted was substantially different between the two academic advisors responsible for the treatment. As such, it became necessary to treat each academic advisor as a separate treatment. The analysis of propensity score models for more than two groups (i.e. two treatments and one control) has relied on conducting three separate analyses. We outline here an approach to conducting propensity score analysis with three groups. 

```{r loaddata, results='hide'}
data(students)
names(students)
```

We will create a `treat` variable that identifies our three groups.

```{r treatsetup}
treat <- students$TreatBy
table(treat, useNA='ifany')
describeBy(students$CreditsAttempted, group=list(students$TreatBy), mat=TRUE, skew=FALSE)
```

The following boxplot shows unadjusted results.

```{r boxplot}
ggplot(students, aes(x=TreatBy, y=CreditsAttempted, colour=TreatBy)) + 
	geom_boxplot() + coord_flip() + geom_jitter()
```

#### Estimate Propensity Scores

The `triangle.psa` function will estimate three propensity score models.

```{r trianglepsa, results='hide', message=FALSE, cache=TRUE}
cols.model <- c('Military','Income', 'Employment', 'NativeEnglish','EdLevelMother',
				'EdLevelFather','HasAssocAtEnrollment','Ethnicity','Gender','Age')
tpsa <- triangle.psa(students[,cols.model], treat, ids=1:nrow(students))
```


```{r psaestimates}
head(tpsa)
(p <- plot(tpsa))
```

#### Matched Triplets

```{r trianglematch, results='hide', message=FALSE, cache=TRUE}
tmatch <- triangle.match(tpsa)
```

Triangle plot of the results. We can see how the propensity scores translate from one model to another.

```{r matches, message=FALSE, cache=FALSE}
head(tmatch)
plot(tmatch, rows=c(2), line.alpha=1, draw.segments=TRUE)
```

We can plot the distances. We can specify other calipers to see how may matched triplets we eliminate if we specify a small caliper to the `triangle.match` function.

```{r distances}
plot.distances(tmatch, caliper=c(.15, .2, .25))
```

The numbers on the left edge are the row numbers from `tmatch`. We can then use the `plot.triangle.matches` function with specifying the `rows` parameters to any or all of these values to investigate that matched triplet. The following figures shows that the large distances in due to the fact that only one data point has a very large propensity score in both model 1 and 2.

```{r followup}
tmatch[tmatch$Dtotal > .11,]
tmatch[838,]
plot(tmatch, rows=c(838), line.alpha=1, draw.segments=TRUE)
```

#### Examine unmatched students.

```{r unmatched}
#Look at the subjects that could not be matched
unmatched <- attr(tmatch, 'unmatched')
nrow(unmatched) / nrow(tpsa) * 100
#Percentage of each group not matched
table(unmatched$treat) / table(tpsa$treat) * 100 
unmatched[unmatched$treat != 'Control',]
```

We can create a triangle plot of only the unmatched students by subsetting `tpsa` with those students in the `unmatched` data frame.

```{r plotunmatched}
plot(tpsa[tpsa$id %in% unmatched$id,])
```

#### Parrellel Plot

```{r merge}
tmatch.out <- merge(x=tmatch, y=students[,c('CreditsAttempted')])
names(tmatch.out)
plot.parallel(tmatch.out)
```

#### Friedman Rank Sum Test

```{r freidman}
outcomes <- grep(".out$", names(tmatch.out), perl=TRUE)
tmatch.out$id <- 1:nrow(tmatch.out)
out <- melt(tmatch.out[,c(outcomes, ncol(tmatch.out))],id.vars='id')
names(out) <- c('ID','Treatment','Outcome')
head(out)
set.seed(2112)
friedman.test(Outcome ~ Treatment | ID, out)
```

#### Repeated Measures ANOVA

```{r repeatedANOVA}
rmanova <- ezANOVA(data=out, dv=Outcome, wid=ID, within=Treatment)
print(rmanova)
```

#### Pairwise Wilcoxon Rank Sum Tests

```{r wilcox}
pairwise.wilcox.test(x=out$Outcome, g=out$Treatment, paired=TRUE, p.adjust.method='bonferroni')
```

#### Posthoc *t*-tests

```{r ttests}
(t1 <- t.test(x=tmatch.out$Treatment1.out, y=tmatch.out$Control.out, paired=TRUE))
(t2 <- t.test(x=tmatch.out$Treatment2.out, y=tmatch.out$Control.out, paired=TRUE))
(t3 <- t.test(x=tmatch.out$Treatment2.out, y=tmatch.out$Treatment1.out, paired=TRUE))
```

#### Boxplot of differences

```{r boxplotdiff}
tmatch.out$Treat1_Control <- tmatch.out$Treatment1.out - tmatch.out$Control.out
tmatch.out$Treat2_Control <- tmatch.out$Treatment2.out - tmatch.out$Control.out
tmatch.out$Treat2_Treat1 <- tmatch.out$Treatment2.out - tmatch.out$Treatment1.out
out.box <- melt(tmatch.out[,c('id','Treat1_Control','Treat2_Control','Treat2_Treat1')], id.vars='id')
names(out.box) <- c('Student','Treatment','Difference')

ci <- as.data.frame(rbind(t1$conf.int, t2$conf.int, t3$conf.int))
ci$Treatment <- names(tmatch.out)[12:14]
ci$estimate <- c(t1$estimate, t2$estimate, t3$estimate)

ggplot(out.box, aes(x=Treatment, y=Difference)) + 
	geom_boxplot() + 
	geom_hline(yintercept=0) +
	geom_crossbar(data=ci, aes(x=Treatment, ymin=V1, ymax=V2, y=estimate), 
				  color='green', fill='green', width=.72, alpha=.6) +
	scale_x_discrete(NULL, labels=c(Treat1_Control='Treat1 - Control', 
									Treat2_Control='Treat2 - Control',
									Treat2_Treat1='Treat2 - Treat1')) +
	xlab(NULL)
```

#### References

Rosenbaum, P.R., & Rubin, D.B. (1983). ￼The Central Role of the Propensity Score in Observational Studies for Causal Effects. *Biometrika, 70*(1).

Rosenbaum, P.R., & Rubin, D.B. (1885) Constructing a control group using multivariate matched sampling methods that incorporate teh propensity score. *The American Statistician, 39*(1).

[National Medical Expenditure Survey](http://dx.doi.org/10.3886/ICPSR09280.v1)

National Center for Health Services Research and Health Care Technology Assessment. NATIONAL MEDICAL EXPENDITURE SURVEY, 1987: INSTITUTIONAL POPULATION COMPONENT. Rockville, MD: Westat, Inc. [producer], 1987. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 1990. doi:10.3886/ICPSR09280.v1
