---
layout: mathjax	
title: TriMatch
subtitle: Propensity Score Matching for Non-Binary Treatments
published: false
status: process
submenu: trimatch
---

#### Example One: New Student Outreach

```{r, results='hide', echo=FALSE}
opts_chunk$set(comment=NA, fig.width=11, fig.path='trimatch/')
options(width=90)
require(TriMatch)
```

The data in example one (`data(students)`) represent newly enrolled students in a distance education program. By the nature of the program all students are considered part-time. Moreover, enrollment in the institution does not necessarily mean students are making active progress towards degree completion. To address this issue the institution began an outreach program whereby academic advisors would regularly contact new students within the first six months of enrollment until either six months have passed, or the student enrolled in some credit earning activity (generally a course or an examination for credit). Since randomization to receive the outreach was not possible, a comparison group was defined as students who enrolled six months prior to the start of the outreach. The treatment group is identified as students who enrolled six months after the start of the outreach and who received at least one academic advisor contact. 

Covariates for estimating propensity scores were retrieved from the student information system. The dependent variable of interest is the number of credits attempted within the first seven months of enrollment.

During the implementation phase it was identified that the outreach being conducted was substantially different between the two academic advisors responsible for the treatment. As such, it became necessary to treat each academic advisor as a separate treatment. The analysis of propensity score models for more than two groups (i.e. two treatments and one control) has relied on conducting three separate analyses. We outline here an approach to conducting propensity score analysis with three groups. 

```{r loaddata, results='hide'}
data(students)
names(students)
```

We will create a `treat` variable that identifies our three groups.

```{r treatsetup}
treat <- students$TreatBy
table(treat, useNA='ifany')
describeBy(students$CreditsAttempted, group=list(students$TreatBy), mat=TRUE, skew=FALSE)
```

The following boxplot shows unadjusted results.

```{r boxplot}
ggplot(students, aes(x=TreatBy, y=CreditsAttempted, colour=TreatBy)) + 
	geom_boxplot() + coord_flip() + geom_jitter()
```

#### Estimate Propensity Scores

The `trips` function will estimate three propensity score models.

```{r trips, results='hide', message=FALSE, cache=TRUE}
cols.model <- c('Military','Income', 'Employment', 'NativeEnglish','EdLevelMother',
				'EdLevelFather','HasAssocAtEnrollment','Ethnicity','Gender','Age')
tpsa <- trips(students[,cols.model], treat, ids=1:nrow(students))
```


```{r psaestimates}
head(tpsa)
(p <- plot(tpsa))
```

#### Matched Triplets

```{r trianglematch, results='hide', message=FALSE, cache=TRUE}
tmatch <- trimatch(tpsa)
```

Triangle plot of the results. We can see how the propensity scores translate from one model to another.

```{r matches, message=FALSE, cache=FALSE}
head(tmatch)
plot(tmatch, rows=c(2), line.alpha=1, draw.segments=TRUE)
```

We can plot the distances. We can specify other calipers to see how may matched triplets we eliminate if we specify a small caliper to the `trimatch` function.

```{r distances}
plot.distances(tmatch, caliper=c(.15, .2, .25))
```

The numbers on the left edge are the row numbers from `tmatch`. We can then use the `plot.triangle.matches` function with specifying the `rows` parameters to any or all of these values to investigate that matched triplet. The following figures shows that the large distances in due to the fact that only one data point has a very large propensity score in both model 1 and 2.

```{r followup}
tmatch[tmatch$Dtotal > .11,]
tmatch[838,]
plot(tmatch, rows=c(838), line.alpha=1, draw.segments=TRUE)
```

#### Examine unmatched students.

```{r unmatched}
#Look at the subjects that could not be matched
unmatched <- attr(tmatch, 'unmatched')
nrow(unmatched) / nrow(tpsa) * 100
#Percentage of each group not matched
table(unmatched$treat) / table(tpsa$treat) * 100 
unmatched[unmatched$treat != 'Control',]
```

We can create a triangle plot of only the unmatched students by subsetting `tpsa` with those students in the `unmatched` data frame.

```{r plotunmatched}
plot(tpsa[tpsa$id %in% unmatched$id,])
```

#### Parrellel Plot

```{r merge}
tmatch.out <- merge(x=tmatch, y=students[,c('CreditsAttempted')])
names(tmatch.out)
plot.parallel(tmatch.out)
```

#### Friedman Rank Sum Test

```{r freidman}
outcomes <- grep(".out$", names(tmatch.out), perl=TRUE)
tmatch.out$id <- 1:nrow(tmatch.out)
out <- melt(tmatch.out[,c(outcomes, ncol(tmatch.out))],id.vars='id')
names(out) <- c('ID','Treatment','Outcome')
head(out)
set.seed(2112)
friedman.test(Outcome ~ Treatment | ID, out)
```

#### Repeated Measures ANOVA

```{r repeatedANOVA}
rmanova <- ezANOVA(data=out, dv=Outcome, wid=ID, within=Treatment)
print(rmanova)
```

#### Pairwise Wilcoxon Rank Sum Tests

```{r wilcox}
pairwise.wilcox.test(x=out$Outcome, g=out$Treatment, paired=TRUE, p.adjust.method='bonferroni')
```

#### Posthoc *t*-tests

```{r ttests}
(t1 <- t.test(x=tmatch.out$Treatment1.out, y=tmatch.out$Control.out, paired=TRUE))
(t2 <- t.test(x=tmatch.out$Treatment2.out, y=tmatch.out$Control.out, paired=TRUE))
(t3 <- t.test(x=tmatch.out$Treatment2.out, y=tmatch.out$Treatment1.out, paired=TRUE))
```

#### Boxplot of differences

```{r boxplotdiff}
tmatch.out$Treat1_Control <- tmatch.out$Treatment1.out - tmatch.out$Control.out
tmatch.out$Treat2_Control <- tmatch.out$Treatment2.out - tmatch.out$Control.out
tmatch.out$Treat2_Treat1 <- tmatch.out$Treatment2.out - tmatch.out$Treatment1.out
out.box <- melt(tmatch.out[,c('id','Treat1_Control','Treat2_Control','Treat2_Treat1')], id.vars='id')
names(out.box) <- c('Student','Treatment','Difference')

ci <- as.data.frame(rbind(t1$conf.int, t2$conf.int, t3$conf.int))
ci$Treatment <- names(tmatch.out)[12:14]
ci$estimate <- c(t1$estimate, t2$estimate, t3$estimate)

ggplot(out.box, aes(x=Treatment, y=Difference)) + 
	geom_boxplot() + 
	geom_hline(yintercept=0) +
	geom_crossbar(data=ci, aes(x=Treatment, ymin=V1, ymax=V2, y=estimate), 
				  color='green', fill='green', width=.72, alpha=.6) +
	scale_x_discrete(NULL, labels=c(Treat1_Control='Treat1 - Control', 
									Treat2_Control='Treat2 - Control',
									Treat2_Treat1='Treat2 - Treat1')) +
	xlab(NULL)
```
