---
layout: default	
title: likert
subtitle: An R package analyzing and visualizing Likert items
published: false
status: process
submenu: none
---

```{r, results='hide', echo=FALSE}
options(width=110)
```

### Overview

The latest version of the `likert` package can be installed from [Github](http://github.com/jbryer/likert) using the `devtools` package:

	require(devtools)
	install_github('likert', 'jbryer')

The following will analyze the reading attitude items from the [Programme of International Student Assessment](http://www.pisa.oecd.org).


```{r setup, message=FALSE}
require(likert)
ls('package:likert')
```

#### Item 28: Reading Attitudes

```{r pisaSetup}
data(pisaitems)

items28 <- pisaitems[,substr(names(pisaitems), 1,5) == 'ST24Q']
head(items28); ncol(items28)

items28 <- rename(items28, c(
			ST24Q01="I read only if I have to.",
			ST24Q02="Reading is one of my favorite hobbies.",
			ST24Q03="I like talking about books with other people.",
			ST24Q04="I find it hard to finish books.",
			ST24Q05="I feel happy if I receive a book as a present.",
			ST24Q06="For me, reading is a waste of time.",
			ST24Q07="I enjoy going to a bookstore or a library.",
			ST24Q08="I read only to get information that I need.",
			ST24Q09="I cannot sit still and read for more than a few minutes.",
			ST24Q10="I like to express my opinions about books I have read.",
			ST24Q11="I like to exchange books with my friends"))
```

```{r likertSummary}
l28 <- likert(items28)
summary(l28)
```

Some plots...

```{r barchart, fig.width=10, fig.path='likert/'}
plot(l28)
plot(l28, centered=TRUE, wrap=30)
plot(l28, type='density')
plot(l28, type='heat')
```

Grouped by country...

```{r likertGrouped, fig.path='likert/'}
l28g <- likert(items28, grouping=pisaitems$CNT)
print(l28g)
summary(l28g)
```

```{r centeredPlot, message=FALSE, results='hide', fig.path='likert/'}
plot(l28g)
plot(l28g, centered=TRUE)
plot(l28g, type='density')
```

#### Item 29: How often do you read these materials because you want to?

```{r item29setup}
title <- "How often do you read these materials because you want to?"
items29 <- pisaitems[,substr(names(pisaitems), 1,5) == 'ST25Q']
names(items29) <- c("Magazines", "Comic books", "Fiction", "Non-fiction books", "Newspapers")
l29 <- likert(items29)
summary(l29)
```

Some plots...

```{r barchart29, fig.width=10, fig.path='likert/'}
plot(l29) + ggtitle(title)
plot(l29, centered=TRUE) + ggtitle(title)
plot(l29, centered=TRUE, center=2.5) + ggtitle(title)
```

Grouped by Country...

```{r item29gsetup, fig.width=10, fig.path='likert/'}
l29g <- likert(items29, grouping=pisaitems$CNT)
summary(l29g)
```

```{r barchart29g, fig.width=10, fig.path='likert/'}
plot(l29g) + ggtitle(title)
plot(l29g, centered=TRUE, center=2.5) + ggtitle(title)
plot(l29g, type='density', legend='Country') + ggtitle(title)
```

