--- 
layout: post
title: str Implementation for Data Frames
tags: R R-Bloggers
type: post
published: true
status: process
---

The `str` function is perhaps the most useful function in R. It provides great information about the structure of some object. When I teach R, especially for those coming from SPSS, the `str` function for data frames provides the information they are use to seeing on the variable view tab. However, sometimes I want to display the information `str` returns in a better format (e.g. as an HTML or LaTeX table). I wrote a function, `strtable` that provides the information `str.data.frame` does but returns the results as a `data.frame`. This provides much more flexibility for controlling how the output is formatted. Specifically, it will return a `data.frame` with four columns: `variable`, `class`, `levels`, and `examples`.

The function can be sourced from [Gist](https://gist.github.com/jbryer/4a0a5ab9fe7e1cf3be0e) using the `devtools` package.

```{r setup, eval=TRUE, results='hide', message=FALSE}
devtools::source_gist('4a0a5ab9fe7e1cf3be0e')
```

For the first example, we'll use the `iris` data frame.

```{r iris}
data(iris)
str(iris)
```

The `strtable` has five parameters:

* `n` the first n element to show
* `width` maximum width in characters for the examples to show
* `n.levels` the first n levels of a factor to show.
* `width.levels` maximum width in characters for the number of levels to show.
* `factor.values` function defining how factor examples should be printed. Possible values are `as.character` or `as.integer`.

```{r irisstrtable}
print(strtable(iris), na.print='')
print(strtable(iris, factor.values=as.integer), na.print='')
```


Here's a second example using the `diamonds` data from the `ggplot2` package.

```{r diamonds}
data(diamonds)
str(diamonds)
print(strtable(diamonds), na.print='')
print(strtable(diamonds, factor.values=as.integer), na.print='')
```

Here's the source code from [Gist](https://gist.github.com/jbryer/4a0a5ab9fe7e1cf3be0e):

<script src="https://gist.github.com/jbryer/4a0a5ab9fe7e1cf3be0e.js"></script>

